Follow-up Samples

### 2017-01-16 submission/manuscript_outline_20161024.Rmd
streamline results

**Background:** Colorectal cancer (CRC) continues to be a worldwide health problem with early detection being used as a key component in mitigating deaths due to the disease. Previous research suggests a link between fecal bacterial microbiome and CRC.   The overall objective was to investigate the changes in the bacterial microbiome after surgery in patients with lesion (i.e. adenoma or carcinoma).  Specifically, we wanted to identify what within the community was different within those undergoing surgical removal of lesion.  We also investigated the use of the bacterial microbiome and Fecal Immunoglobulin Test (FIT) to build models to either classify individuals as having a lesion or whether, based on the bacterial microbiome, the sample could be classified correctly as before or after surgery.   
**Results:** Adenoma individual's bacterial microbiome were more similar to their pre-surgery sample then those with carcinoma (P-value = `r format(change_summary['thetayc', 'Pvalue'], digits = 3)`) and this was also reflected in FIT  (P-value = `r format(change_summary['fit', 'Pvalue'], scientific = TRUE, digits = 3)`).  There was no significant difference in any indivdiual OTU between samples before and after surgery (P-value > `r format(min(filter(all_OTU_pair_wilcox, analysis == "ALL")[, 'BH_corrected']), digits = 3)`).  A model with a total of `r length(lesion_top10_varsMDA$variable)` variables was able to classify lesion (AUC = `r format(max(reduced_lesion_AUC_run_summary$ROC), digits = 3)` - `r format(min(reduced_lesion_AUC_run_summary$ROC), digits = 3)` while the model to classify samples as before and after surgery had `r length(IF_top10_varsMDA$variable)` variables (AUC = `r format(max(reduced_IF_AUC_run_summary$ROC), digits = 3)` - `r format(min(reduced_IF_AUC_run_summary$ROC), digits = 3)`).  Both models had a significant decrease in the positive probability of a lesion or being a before surgery sample (P-value = `r format(filter(all_probs_paired_wilcox, model_type == "red_lesion" & comparison =="lesion")[, "BH_correction"], scientific = TRUE, digits = 3)` and `r format(filter(all_probs_paired_wilcox, model_type == "red_IF" & comparison =="lesion")[, "BH_correction"], scientific = TRUE, digits = 3)`).  In total there were `r length(common_vars_summary$otu)` OTUs that were common to both models and all of these classified to commensals with the largest representation belonging to `r common_vars_summary["Otu000005", "lowest_ID"]`, `r common_vars_summary["Otu000012", "lowest_ID"]`, `r common_vars_summary["Otu000082", "lowest_ID"]`, and `r paste(gsub("_unclassified", "", common_vars_summary["Otu000006", "lowest_ID"]))`.   
**Conclusions:** Our data suggests that treatment not only significantly reduces the probability of having a colonic lesion but also causes detectable changes in the bacterial microbiome.  Further surveillance of these individuals will enable us to determine whether models such as the one we present here can also be used to predict recurrence of colorectal cancer.

@@

***Bacterial Community and Fit Changes before and after Treatment***
Based on thetayc distance metrics, comparing the initial to the follow up samples, there was a significant difference between the adenoma and carcinoma groups (P-value = `r format(change_summary['thetayc', 'Pvalue'], scientific = FALSE, digits = 3)`) [Figure 1a]. There was also a significant difference in FIT between initial and follow up samples with the carcinoma group having a significant decrease in FIT versus the adenoma group (P-value = `r format(change_summary['fit', 'Pvalue'], scientific = TRUE, digits = 3)`) [Figure 1b].  The whole community structure before and after surgery were visualized on NMDS graphs for both adenoma [Figure1c] (PERMANOVA = `r format(beta_summary['Carcinoma', 'PERMANOVA'], scientific = FALSE, digits = 3)`) and carcinoma [Figure 1d] (PERMANOVA = `r format(beta_summary['Adenoma', 'PERMANOVA'], scientific = FALSE, digits = 3)`).  When all initial and follow up samples were compared to each other there was no significant overall difference between them (PERMANOVA = `r format(beta_summary['All', 'PERMANOVA'], scientific = FALSE, digits = 3)`).  There was also no significant difference between initial and follow up samples for observed OTUs, Shannon diversity, or evenness after correction for multiple comparisons [Table S1].  There was no significant difference between initial and follow up samples for any single OTU [Figure S1].

***Differences in Adenoma and Carcinoma of Previously Associated Cancer Bacteria***
We next examined whether there were differences in previously well described carcinoma associated OTUs.  These included the OTUs that aligned with Porphyromonas asaccharolytica (`r unique(filter(crc_bug_info, tax_id == "Porphyromonas")[, "otu"])`), Fusobacterium nucleatum (`r unique(filter(crc_bug_info, tax_id == "Fusobacterium")[, "otu"])`), Parvimonas micra (`r unique(filter(crc_bug_info, tax_id == "Parvimonas")[, "otu"])`), and Peptostreptococcus stomatis (`r unique(filter(crc_bug_info, tax_id == "Peptostreptococcus")[, "otu"])`).  First, the carcinoma samples showed a significant difference between initial and follow up samples for Peptostreptococcus stomatis (P-value = `r format(main_crc_bugs_summary["pept", "crc_BH"], digits = 3)`) and Porphyromonas asaccharolytica (P-value = `r format(main_crc_bugs_summary["porp", "crc_BH"], digits = 3)`) whereas there were no signficant differences in any of these OTUs in the adenoma samples [Table S4]. Second, when these OTUs were present, there was a clear magnitude difference in these specific OTUs based on whether they were from adenoma or carcinoma individuals [Figure 2].  However, only a small percentage of those with adenoma or carcinoma were positive for any of these OTUs.

***Full and Reduced Model Results***
Since differences were observed between initial and follow up samples and only a small number of individuals were positive for previously associated CRC bacteria we next investigated if we could create a model that could adequately classify and adjust lesion probability based on the bacterial community and FIT.  The lesion model had an AUC range of `r format(min(lesion_AUC_run_summary$ROC), digits = 3)` to `r format(max(lesion_AUC_run_summary$ROC), digits = 3)` versus the initial follow up model which had and AUC range of `r format(min(IF_AUC_run_summary$ROC), digits = 3)` to `r format(max(IF_AUC_run_summary$ROC), digits = 3)` after 100 20 repeated 10-fold cross validations.  Interestingly, identification of the most important variables and reducing the models to only these factors increased the AUC in the lesion model (`r format(min(reduced_lesion_AUC_run_summary$ROC), digits = 3)` - `r format(max(reduced_lesion_AUC_run_summary$ROC), digits = 3)`) and initial and follow up model (`r format(min(reduced_IF_AUC_run_summary$ROC), digits = 3)` - `r format(max(reduced_IF_AUC_run_summary$ROC), digits = 3)`).

The test set AUC range for the full and reduced lesion model were similar to that reported for the training set AUC ranges and the ROC curve ranges overlap each other [Figure 3a].  The ROC curve for the final lesion model used falls within the range of both the full and reduced lesion model [Figure 3a].  Interestingly, the test set AUC range for the initial and follow up performed much better then the training set AUCs.  Both the full and reduced initial and follow up models overlapped with each other [Figure 3b] there was a marked decrease in the ROC curve for the final model used.

>>>>

**Background:** Colorectal cancer (CRC) continues to be a worldwide health problem with early detection being used as a key component in mitigating deaths due to the disease. Previous research suggests a link between fecal bacterial microbiome and CRC.   The overall objective was to investigate the changes in the bacterial microbiome after surgery in patients with lesion (i.e. adenoma or carcinoma).  Specifically, we wanted to identify what within the community was different within those undergoing surgical removal of lesion.  We also investigated the use of the bacterial microbiome and Fecal Immunoglobulin Test (FIT) to build models to either classify individuals as having a lesion or whether, based on the bacterial microbiome, the sample could be classified correctly as before or after surgery.   
**Results:** Adenoma individual's bacterial microbiome were more similar to their pre-surgery sample then those with carcinoma (P-value = `r format(change_summary['thetayc', 'Pvalue'], digits = 3)`) and this was also reflected in FIT  (P-value = `r format(change_summary['fit', 'Pvalue'], scientific = TRUE, digits = 3)`).  There was no significant difference in any indivdiual OTU between samples before and after surgery (P-value > `r format(min(filter(all_OTU_pair_wilcox, analysis == "ALL")[, 'BH_corrected']), digits = 3)`).  A model with a total of `r length(lesion_top10_varsMDA$variable)` variables was able to classify lesion (AUC = `r format(max(reduced_lesion_AUC_run_summary$ROC), digits = 3)` - `r format(min(reduced_lesion_AUC_run_summary$ROC), digits = 3)` while the model to classify samples as before and after surgery had `r length(IF_top10_varsMDA$variable)` variables (AUC = `r format(max(reduced_IF_AUC_run_summary$ROC), digits = 3)` - `r format(min(reduced_IF_AUC_run_summary$ROC), digits = 3)`).  Both models had a significant decrease in the positive probability of a lesion or being a before surgery sample (P-value = `r format(filter(all_probs_paired_wilcox, model_type == "red_lesion" & comparison =="lesion")[, "BH_correction"], scientific = TRUE, digits = 3)` and `r format(filter(all_probs_paired_wilcox, model_type == "red_IF" & comparison =="lesion")[, "BH_correction"], scientific = TRUE, digits = 3)`).  In total there were `r length(common_vars_summary$otu)` OTUs that were common to both models and the majority of these classified to commensal bacteria with the largest representation belonging to `r common_vars_summary["Otu000005", "lowest_ID"]`, `r common_vars_summary["Otu000012", "lowest_ID"]`, `r common_vars_summary["Otu000082", "lowest_ID"]`, and `r paste(gsub("_unclassified", "", common_vars_summary["Otu000006", "lowest_ID"]))`.   
**Conclusions:** Our data suggests that treatment not only significantly reduces the probability of having a colonic lesion but also causes detectable changes in the bacterial microbiome.  Further surveillance of these individuals will enable us to determine whether models such as the one we present here can also be used to predict recurrence of colorectal cancer.

@@

***Bacterial Community and FIT Changes before and after Surgery***
Based on thetayc distance metrics, comparing the initial to the follow up samples, there was a significant difference between the adenoma and carcinoma groups (P-value = `r format(change_summary['thetayc', 'Pvalue'], scientific = FALSE, digits = 3)`) [Figure 1a]. There was also a significant difference in change in FIT between initial and follow up samples (P-value = `r format(change_summary['fit', 'Pvalue'], scientific = TRUE, digits = 3)`) [Figure 1b].  The whole community structure before and after surgery were visualized on NMDS graphs for both adenoma [Figure1c] (PERMANOVA = `r format(beta_summary['Adenoma', 'PERMANOVA'], scientific = FALSE, digits = 3)`) and carcinoma [Figure 1d] (PERMANOVA = `r format(beta_summary['Carcinoma', 'PERMANOVA'], scientific = FALSE, digits = 3)`).  When all initial and follow up samples were compared to each other there was no significant overall difference between them (PERMANOVA = `r format(beta_summary['All', 'PERMANOVA'], scientific = FALSE, digits = 3)`).  There was no significant difference between initial and follow up samples for observed OTUs, Shannon diversity, or evenness after correction for multiple comparisons [Table S1].  There was also no significant difference between initial and follow up samples for any single OTU [Figure S1].

***Differences in Adenoma and Carcinoma of Previously Associated Cancer Bacteria***
We next examined whether there were differences in previously well described carcinoma associated OTUs.  These included the OTUs that aligned with *Porphyromonas asaccharolytica* (`r unique(filter(crc_bug_info, tax_id == "Porphyromonas")[, "otu"])`), *Fusobacterium nucleatum* (`r unique(filter(crc_bug_info, tax_id == "Fusobacterium")[, "otu"])`), *Parvimonas micra* (`r unique(filter(crc_bug_info, tax_id == "Parvimonas")[, "otu"])`), and *Peptostreptococcus stomatis* (`r unique(filter(crc_bug_info, tax_id == "Peptostreptococcus")[, "otu"])`).  First, the carcinoma samples showed a significant difference between initial and follow up samples for *Peptostreptococcus stomatis* (P-value = `r format(main_crc_bugs_summary["pept", "crc_BH"], digits = 3)`) and *Porphyromonas asaccharolytica* (P-value = `r format(main_crc_bugs_summary["porp", "crc_BH"], digits = 3)`) whereas there was no signficant differences in any of these OTUs in the adenoma samples [Table S4]. Second, when these OTUs were present, there was a clear magnitude difference based on whether they were from adenoma or carcinoma [Figure 2].  However, only a small percentage of those with adenoma or carcinoma were positive for any of these OTUs.

***Full and Reduced Model Results***
Since differences were observed between initial and follow up samples and only a small number of individuals were positive for previously associated CRC bacteria; we next investigated if we could create a model that could adequately classify and adjust lesion probability based on the bacterial community and FIT.  The lesion model had an AUC range of `r format(min(lesion_AUC_run_summary$ROC), digits = 3)` to `r format(max(lesion_AUC_run_summary$ROC), digits = 3)` versus the initial follow up model which had an AUC range of `r format(min(IF_AUC_run_summary$ROC), digits = 3)` to `r format(max(IF_AUC_run_summary$ROC), digits = 3)` after 100 20 repeated 10-fold cross validations.  Interestingly, identification of the most important variables and reducing the models to only these factors increased the AUC in the lesion model (`r format(min(reduced_lesion_AUC_run_summary$ROC), digits = 3)` - `r format(max(reduced_lesion_AUC_run_summary$ROC), digits = 3)`) and initial and follow up model (`r format(min(reduced_IF_AUC_run_summary$ROC), digits = 3)` - `r format(max(reduced_IF_AUC_run_summary$ROC), digits = 3)`).

The test set AUC range for the full and reduced lesion model were similar to that reported for the training set AUC ranges and the ROC curve ranges overlap each other [Figure 3a].  The ROC curve for the final lesion model used was within the range of both the full and reduced lesion model [Figure 3a].  Interestingly, the test set AUC range for the initial and follow up performed much better then the training set AUCs.  Both the full and reduced initial and follow up models overlapped with each other [Figure 3b] there was a marked decrease in the ROC curve for the final model used.

### 2017-01-16 submission/manuscript_outline_20161024.tex
streamline results

\textbf{Background:} Colorectal cancer (CRC) continues to be a worldwide
health problem with early detection being used as a key component in
mitigating deaths due to the disease. Previous research suggests a link
between fecal bacterial microbiome and CRC. The overall objective was to
investigate the changes in the bacterial microbiome after surgery in
patients with lesion (i.e.~adenoma or carcinoma). Specifically, we
wanted to identify what within the community was different within those
undergoing surgical removal of lesion. We also investigated the use of
the bacterial microbiome and Fecal Immunoglobulin Test (FIT) to build
models to either classify individuals as having a lesion or whether,
based on the bacterial microbiome, the sample could be classified
correctly as before or after surgery.\\
\textbf{Results:} Adenoma individual's bacterial microbiome were more
similar to their pre-surgery sample then those with carcinoma (P-value =
0.00198) and this was also reflected in FIT (P-value = 2.15e-05). There
was no significant difference in any indivdiual OTU between samples
before and after surgery (P-value \textgreater{} 0.125). A model with a
total of 37 variables was able to classify lesion (AUC = 0.847 - 0.791
while the model to classify samples as before and after surgery had 33
variables (AUC = 0.79 - 0.651). Both models had a significant decrease
in the positive probability of a lesion or being a before surgery sample
(P-value = 1.91e-11 and 6.72e-12). In total there were 14 OTUs that were
common to both models and all of these classified to commensals with the
largest representation belonging to Bacteroides, Blautia, Streptococcus,
and Clostridiales.\\
\textbf{Conclusions:} Our data suggests that treatment not only
significantly reduces the probability of having a colonic lesion but
also causes detectable changes in the bacterial microbiome. Further
surveillance of these individuals will enable us to determine whether
models such as the one we present here can also be used to predict
recurrence of colorectal cancer.

@@

\textbf{\emph{Bacterial Community and Fit Changes before and after
Treatment}} Based on thetayc distance metrics, comparing the initial to
the follow up samples, there was a significant difference between the
adenoma and carcinoma groups (P-value = 0.00198) {[}Figure 1a{]}. There
was also a significant difference in FIT between initial and follow up
samples with the carcinoma group having a significant decrease in FIT
versus the adenoma group (P-value = 2.15e-05) {[}Figure 1b{]}. The whole
community structure before and after surgery were visualized on NMDS
graphs for both adenoma {[}Figure1c{]} (PERMANOVA = 0.002) and carcinoma
{[}Figure 1d{]} (PERMANOVA = 0.997). When all initial and follow up
samples were compared to each other there was no significant overall
difference between them (PERMANOVA = 0.085). There was also no
significant difference between initial and follow up samples for
observed OTUs, Shannon diversity, or evenness after correction for
multiple comparisons {[}Table S1{]}. There was no significant difference
between initial and follow up samples for any single OTU {[}Figure
S1{]}.

\textbf{\emph{Differences in Adenoma and Carcinoma of Previously
Associated Cancer Bacteria}} We next examined whether there were
differences in previously well described carcinoma associated OTUs.
These included the OTUs that aligned with Porphyromonas asaccharolytica
(Otu000153), Fusobacterium nucleatum (Otu000226), Parvimonas micra
(Otu000460), and Peptostreptococcus stomatis (Otu000653). First, the
carcinoma samples showed a significant difference between initial and
follow up samples for Peptostreptococcus stomatis (P-value = 0.0183) and
Porphyromonas asaccharolytica (P-value = 0.0154) whereas there were no
signficant differences in any of these OTUs in the adenoma samples
{[}Table S4{]}. Second, when these OTUs were present, there was a clear
magnitude difference in these specific OTUs based on whether they were
from adenoma or carcinoma individuals {[}Figure 2{]}. However, only a
small percentage of those with adenoma or carcinoma were positive for
any of these OTUs.

\textbf{\emph{Full and Reduced Model Results}} Since differences were
observed between initial and follow up samples and only a small number
of individuals were positive for previously associated CRC bacteria we
next investigated if we could create a model that could adequately
classify and adjust lesion probability based on the bacterial community
and FIT. The lesion model had an AUC range of 0.723 to 0.795 versus the
initial follow up model which had and AUC range of 0.451 to 0.67 after
100 20 repeated 10-fold cross validations. Interestingly, identification
of the most important variables and reducing the models to only these
factors increased the AUC in the lesion model (0.791 - 0.847) and
initial and follow up model (0.651 - 0.79).

The test set AUC range for the full and reduced lesion model were
similar to that reported for the training set AUC ranges and the ROC
curve ranges overlap each other {[}Figure 3a{]}. The ROC curve for the
final lesion model used falls within the range of both the full and
reduced lesion model {[}Figure 3a{]}. Interestingly, the test set AUC
range for the initial and follow up performed much better then the
training set AUCs. Both the full and reduced initial and follow up
models overlapped with each other {[}Figure 3b{]} there was a marked
decrease in the ROC curve for the final model used.

>>>>

\textbf{Background:} Colorectal cancer (CRC) continues to be a worldwide
health problem with early detection being used as a key component in
mitigating deaths due to the disease. Previous research suggests a link
between fecal bacterial microbiome and CRC. The overall objective was to
investigate the changes in the bacterial microbiome after surgery in
patients with lesion (i.e.~adenoma or carcinoma). Specifically, we
wanted to identify what within the community was different within those
undergoing surgical removal of lesion. We also investigated the use of
the bacterial microbiome and Fecal Immunoglobulin Test (FIT) to build
models to either classify individuals as having a lesion or whether,
based on the bacterial microbiome, the sample could be classified
correctly as before or after surgery.\\
\textbf{Results:} Adenoma individual's bacterial microbiome were more
similar to their pre-surgery sample then those with carcinoma (P-value =
0.00198) and this was also reflected in FIT (P-value = 2.15e-05). There
was no significant difference in any indivdiual OTU between samples
before and after surgery (P-value \textgreater{} 0.125). A model with a
total of 37 variables was able to classify lesion (AUC = 0.847 - 0.791
while the model to classify samples as before and after surgery had 33
variables (AUC = 0.79 - 0.651). Both models had a significant decrease
in the positive probability of a lesion or being a before surgery sample
(P-value = 1.91e-11 and 6.72e-12). In total there were 14 OTUs that were
common to both models and the majority of these classified to commensal
bacteria with the largest representation belonging to Bacteroides,
Blautia, Streptococcus, and Clostridiales.\\
\textbf{Conclusions:} Our data suggests that treatment not only
significantly reduces the probability of having a colonic lesion but
also causes detectable changes in the bacterial microbiome. Further
surveillance of these individuals will enable us to determine whether
models such as the one we present here can also be used to predict
recurrence of colorectal cancer.

@@

\textbf{\emph{Bacterial Community and FIT Changes before and after
Surgery}} Based on thetayc distance metrics, comparing the initial to
the follow up samples, there was a significant difference between the
adenoma and carcinoma groups (P-value = 0.00198) {[}Figure 1a{]}. There
was also a significant difference in change in FIT between initial and
follow up samples (P-value = 2.15e-05) {[}Figure 1b{]}. The whole
community structure before and after surgery were visualized on NMDS
graphs for both adenoma {[}Figure1c{]} (PERMANOVA = 0.997) and carcinoma
{[}Figure 1d{]} (PERMANOVA = 0.002). When all initial and follow up
samples were compared to each other there was no significant overall
difference between them (PERMANOVA = 0.085). There was no significant
difference between initial and follow up samples for observed OTUs,
Shannon diversity, or evenness after correction for multiple comparisons
{[}Table S1{]}. There was also no significant difference between initial
and follow up samples for any single OTU {[}Figure S1{]}.

\textbf{\emph{Differences in Adenoma and Carcinoma of Previously
Associated Cancer Bacteria}} We next examined whether there were
differences in previously well described carcinoma associated OTUs.
These included the OTUs that aligned with \emph{Porphyromonas
asaccharolytica} (Otu000153), \emph{Fusobacterium nucleatum}
(Otu000226), \emph{Parvimonas micra} (Otu000460), and
\emph{Peptostreptococcus stomatis} (Otu000653). First, the carcinoma
samples showed a significant difference between initial and follow up
samples for \emph{Peptostreptococcus stomatis} (P-value = 0.0183) and
\emph{Porphyromonas asaccharolytica} (P-value = 0.0154) whereas there
was no signficant differences in any of these OTUs in the adenoma
samples {[}Table S4{]}. Second, when these OTUs were present, there was
a clear magnitude difference based on whether they were from adenoma or
carcinoma {[}Figure 2{]}. However, only a small percentage of those with
adenoma or carcinoma were positive for any of these OTUs.

\textbf{\emph{Full and Reduced Model Results}} Since differences were
observed between initial and follow up samples and only a small number
of individuals were positive for previously associated CRC bacteria; we
next investigated if we could create a model that could adequately
classify and adjust lesion probability based on the bacterial community
and FIT. The lesion model had an AUC range of 0.723 to 0.795 versus the
initial follow up model which had an AUC range of 0.451 to 0.67 after
100 20 repeated 10-fold cross validations. Interestingly, identification
of the most important variables and reducing the models to only these
factors increased the AUC in the lesion model (0.791 - 0.847) and
initial and follow up model (0.651 - 0.79).

The test set AUC range for the full and reduced lesion model were
similar to that reported for the training set AUC ranges and the ROC
curve ranges overlap each other {[}Figure 3a{]}. The ROC curve for the
final lesion model used was within the range of both the full and
reduced lesion model {[}Figure 3a{]}. Interestingly, the test set AUC
range for the initial and follow up performed much better then the
training set AUCs. Both the full and reduced initial and follow up
models overlapped with each other {[}Figure 3b{]} there was a marked
decrease in the ROC curve for the final model used.
